Layer based routing to obtain quality optimization of scalable video streaming in the Future Internet

### 2012-12-06 paper/paper.tex
Paper

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Low-level routing for quality optimization of scalable video streaming on the future Internet}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Piet Smet, Sander Vrijders, Youri Flement, Bruno Chevalier}
\IEEEauthorblockA{University of Ghent - Computer Science, Ghent, Belgium}
}

@@

\begin{abstract}
%\boldmath
In current IP networks, routes are usually based on the hop count from sender to receiver without keeping in mind the throughput and loss of packets. IP is a best-effort protocol and doesn't guarantee zero packet loss. This is unsatisfactory if we want to deliver qualitative, scalable video streams. This paper presents a framework and algorithms used to provide Quality of Service (\textit{QoS}) support for scalable video streaming in an OpenFlow network. We aim to solve two problems: first we route the base layer of SVC encoded videos as a lossless-QoS flow while the enhancement layers are routed as a lossy-QoS or best effort flow. Second, we provide failure recovery so the OpenFlow network can converge back to an optimal state when links go down. We show that using different costs to calculate paths for QoS flows can improve the overall throughput of video streams when network convergation and delay occurs.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

\section{Introduction}
In a standarized IP network, each router or switch contains a flow table containing the next hop for each destination, often based on the shortest path calculation. The network devices do not know what data a packet contains and they do not care either, each flow follows the same rules. The Internet Protocol (IP) headers contain a priority field but this is often neglected by most routers. Media applications, such as video streaming, require no packet loss and minimal network delay due to their high Quality of Service (QoS) dependancy. When these QoS flows are sent through the same path as other best effort traffic they are subject to network congestion and delays which renders the outcome unsatisfactory.
Scalable Video Coding (SVC) encodes the stream in a base layer along with several enhancement layers. In order to obtain correct output it is important that the base layer has no packet loss and no network delay, something that can't be guaranteed if there are no seperate forwarding rules set up for this QoS flow. To solve this issue we need the routing algorithms to recognize QoS flows and seperate them from normal best effort traffic.\\In this project we are looking to use OpenFlow in order to provide QoS support for scalable video streaming. To achieve this goal we can route the base layer of SVC encoded video as a lossless-QoS flow while the enhancement layers are routed as a lossy-QoS or best effort flow, as described in. Even though this approach deviates from the shortest path algorithms, we can still show that alternative paths can increase the video’s throughput in case of network congestion. We have also provided means of failure recovery so the OpenFlow network c
 an conve
 rge back to an optimal state when links go down.\\OpenFlow [1] is an Internet architecture developed by Stanford University that enables researchers to run experimental protocols in existing networks with heterogeneous switches. This is achieved by decoupling the control and forwarding layers of routing, where each flow is associated with an action. With OpenFlow, vendors do not need to expose the internal working of their products as OpenFlow exploits a common set of functions present in most switches to achieve its goal. Switches forward packages of new flows to a centralized Controller, after which the Controller can set up the routing tables of switches in the network. Listener modules on the Controller can use filters to seperate our QoS flows from other traffic.\\In Section II of this paper we compare related research to our own. In section III we take a closer look at the components of our framework and the principals it is based on. In section IV we present the diffe
 rent alg
 orithms used to determine the behavior of each flow in the network. Section V shows the test scenario we used to evaluate our framework while Section VI presents the found results.

\begin{figure*}[t]
\centering
\includegraphics[width=\linewidth]{SVC_OpenFlow_flow}
\caption{Simulation Results}
\label{fig_sim}
\end{figure*}

\section{Related work}
In our research we identified a need for seperate forwarding actions per flow in order apply load-balancing. The same principal has been discussed in research papers from Princeton University [2]. In their research the aim was to apply load-balancing during service discovery of web services when all requests had to be processed by a central component. A problem occurs when each request is forwarded to a centralized component responsible for service discovery, which leads to network congestion. By using OpenFlow, the researchers at Princeton University managed to set seperate forwarding actions for client requests which match certain criteria. These actions allow forwarding entities to redirect client requests to an appropriate service instance without interference from the centralized component. Even though the use case may be different, the principal is the same as the one applied in our research. When both best effort and QoS flows follow the same paths, chances for congest
 ion and 
 packet loss to occur increase significantly. OpenFlow allows us to set different paths for both flows, which improves the throughput of our QoS flows and also decrease the network load on links used by best effort traffic.\\The idea to provide a minimal quality for SVC encoded videos originates from research described in [4] and [5]. By routing the base layer as a lossless-QoS flow we can guarantee a minimal quality to the user. Ideally, the enhancement layers will also be delivered without packet loss or delay but when congestion occurs we demote the enhancement layers to best effort traffic and prioritise the base layer. Research discussed in [4] also attempts to route the enhancement layers as a lossy-QoS flow if the network allows it.\\The OpenFlow Controller becomes a possible bottleneck for wide-area networks if too many QoS flows need to be processed for a first time. HyperFlow [3] suggests using areas across the network, each containing atleast one Controller. The co
 ntroller
 s of different areas can then exchange information about their areas to obtain a global overview. In our research we work with relatively small networks so we assume that one Controller is sufficient.

\section{Architecture}
Each OpenFlow network contains atleast one Controller which is responsible for managing all the switches, this is no different in our research. In this project we used Floodlight Controller to implement our Java modules, which contain all non-default code. A default OpenFlow switch forwards packages to the Controller if the package does not match any forwarding entry. We study SVC encoded video streams as QoS flow, where we consider the base layer as a lossless-QoS flow and the enhancement layer as best effort traffic, when congestion occurs. These principals are shown in figure 1. A filter on the Controller seperates QoS flows from other traffic, after which the Controller can decide if the shortest path is still sufficient to carry the base layer. When the package is part of a QoS flow and congestion occurs on the shortest path, we calculate an alternative path using metrics other than just the hop count.\\This was one of the first tasks we performed during our research: to
  find me
 trics, besides the hop count, which can be used to calculate the cost of a path to transfer zero packet loss QoS traffic. These metrics are part of a cost function which the routing algorithm will attempt to minimize when calculating optimal paths. We made similar assumptions about QoS-routing as those made in [4] and [5] before defining a cost function:
\begin{itemize}
	\item When lossless-QoS flows are forwarded on the same link as best effort traffic, there will be packet loss in the best effort flow to prioritise the QoS flow.
	\item Metrics for QoS flows must be different than those used for best effort traffic. The shortest path and QoS path must be different to get decent results. If both follow the same path, the best effort traffic will be completely sacrificed for the sake of the lossless-QoS base layer.
	\item Calculating a path for a new QoS flow does not change the best effort routing tables. The packet loss of best effort traffic caused by QoS flows must be considered during path selection.
	\item The length of alternative paths for QoS flows should still be kept as short as possible to optimize transfer time.
	\item Best effort traffic can be used as a reference to retreive metrics from a switch or link, such as packet loss or to estimate the amount of traffic on that route.
\end{itemize}
These assumptions indicate that packet loss and path length are part of our cost function: the cost function must find a balance between maintaining low packet loss of the best effort traffic and the delay of QoS flows, caused by using longer paths.\\Furthermore, the Controller must be able to obtain these metrics at any time from any switch in the network before calculating new flow tables. This is why we implemented a monitor module on the Controller to gather all network statistics. Besides polling for statistics, a switch must also forward events to the Controller when they occur, such as an adjacent link going down or a flow experiencing packet loss. These events provide additional monitoring data for the routing algorithms on the Controller.\\Decisions made by the Controller are sent to the switches as updated forwarding tables. When the shortest path can still ensure zero packet loss, the Dijkstra shortest path algorithm is executed, which is the default implementation
  in Floo
 dlight.

\section{QoS Routing}
Must ensure zero packet loss base layer, optimise other traffic ... cost function?
Go from theoretical cost function to exact function in code?

\section{Evaluation}
Test scenario (client/sever - filter based on port) + metrics for result + present results\\
An evaluation metric has to represent a value which is relevant to what we want to improve, which is why a measurement must:
\begin{itemize}
	\item represent the quality of the received video
	\item represent the performance of the network
	\item represent the impact on best effort traffic
\end{itemize}
To measure quality of the video we have chosen for the PSNR value of the received video compared to the raw footage. The higher the PSNR value, the better the frames were preserved during transfer. To measure the performance of the network we used the bitrate of all traffic. We also show the packet loss of best effort traffic caused by prioritising the QoS flow...

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Layer based routing to obtain quality optimization of scalable video streaming in the Future Internet}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Piet Smet, Sander Vrijders, Youri Flement, Bruno Chevalier}
\IEEEauthorblockA{Ghent University - iMinds, Department of Information Technology,\\ 
Gaston Crommenlaan 8 bus 201, 9050 Gent, Belgium}
}

@@

\begin{abstract}
%\boldmath
In TCP/IP networks, there is no inherent support for Quality of Service.
Routing is done on the minimum hop count from sender to receiver, and all traffic is best-effort traffic. 
This is unsatisfactory for applications such as video streaming. 
This paper presents an OpenFlow controller application which provides 
Quality of Service (\textit{QoS}) support for Scalable Video Streaming (\textit{SVC}) in an OpenFlow network. 
First we route the base layer of SVC encoded videos as a 
high priority QoS flow while the enhancement layers are routed at a lower QoS or as a best-effort flow. 
Second, we provide failure recovery so the OpenFlow network can converge back to an optimal 
state when links fail. We show that using different costs to calculate paths for different flows 
can improve the overall throughput of video streams when network congestion and delay occur.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

\section{Introduction}
In a TCP/IP-based network, each router or switch contains a flow table containing the 
next hop for each destination, based on the shortest path calculation. 
Media applications, such as video streaming, 
require no packet loss and minimal network delay due to their high Quality of Service (QoS) dependency. 
When these QoS flows are sent over the same path as other best effort traffic they 
are subject to network congestion and delay which renders an unsatisfactory outcome. 
Several efforts have been made to support QoS in a TCP/IP-based network, such as DiffServ and IntServ, but these 
solutions only work well in managed networks.
Scalable Video Coding (SVC) encodes the stream in a base layer along with several enhancement layers. 
In order to obtain correct output it is important that the base layer has no 
packet loss and minimal delay, which is not possible without support for QoS. 
To solve this issue we need the routing algorithms to detect QoS flows and route these flows differently than 
normal best effort traffic.
To achieve this goal in video streaming, we route the base layer of SVC encoded video at a high QoS 
while the enhancement layers are routed either at a lower QoS or as a best effort flow. 
We also provide failure recovery so the network 
can converge back to an optimal state when links go down.\\

We use OpenFlow in order to provide QoS support for scalable video streaming. 
OpenFlow is a protocol for switches, allowing to control the packet forwarding process.
This way, networks are not only configurable, but also programmable.
Each OpenFlow Switch has a flow table, which is used for packet lookup and 
forwarding, and a secure channel to an external controller (see Figure \ref{fig:openflow1}). 
The flow table consists of flow entries.
Each flow table entry matches a traffic pattern, by matching the header fields.
The header fields that can be matched in OpenFlow 1.0 can be seen in Table \ref{tab:matchedflows}.
The controller makes the network programmable by managing the switch over 
the secure channel using the OpenFlow protocol.
The controller installs new entries in the flow table to manage the network traffic.
Entries have a hard and a soft timeout. A soft timeout means the entry gets removed if it has not been used for a certain time.
A hard timeout means the flow will be removed after the timeout, even if it is still in use.
There are many implementations of controllers available for OpenFlow switches.
The implementation of the controller we will use is Floodlight, because it is open source and widely used.

\begin{table}[htb]
\centering
\begin{tabular}{| c | *{11}{c|}}
\hline 
In Port & Ethernet Source Address & Ethernet Destination Address \\ \hline
Ethernet Type & VLAN ID & VLAN Priority \\ \hline
IP Source & IP Destination & IP Protocol \\ \hline
IP ToS Bits & TCP/UDP Source Port & TCP/UDP Destination Port\\ \hline 
\end{tabular}
\caption{Fields that can be matched in OpenFlow 1.0}
\label{tab:matchedflows}
\end{table}

\begin{figure}[htb]
\centering
\includegraphics[scale=0.6]{fig/openflow1}
\caption{How OpenFlow works}
\label{fig:openflow1}
\end{figure}

The rest of this paper is organized as follows.
In Section II of this paper we compare related work to our own. 
In section III we take a closer look at the components of our framework 
and the principals it is based on. In section IV we present the different algorithms 
used to determine the behavior of each flow in the network. Section V shows the test 
scenario we used to evaluate our framework while Section VI presents results.
In Section VII we come to a conclusion.

\section{Related work}
In our research we identified a need for separate forwarding actions per flow.
The same principal has been discussed in research papers from Princeton University. 
In their research the aim was to apply load-balancing during service discovery of web services 
when all requests had to be processed by a central component. 
A problem occurs when each request is forwarded to a centralized component responsible 
for service discovery, which leads to network congestion. By using OpenFlow, the researchers at 
Princeton University managed to set separate forwarding actions for client requests which match 
certain criteria. These actions allow forwarding entities to redirect client requests to an 
appropriate service instance without interference from the centralized component. Even though 
the use case may be different, the principal is the same as the one applied in our research. 
When both best effort and QoS flows follow the same paths, chances for congestion and packet loss 
to occur increase significantly. OpenFlow allows us to set different paths for both flows, which 
improves the throughput of our QoS flows and also decrease the network load on links used by best 
effort traffic.\\

The idea to provide a minimal quality for SVC encoded videos originates from 
research described in [4] and [5]. By routing the base layer as a lossless-QoS flow we can 
guarantee a minimal quality to the user. Ideally, the enhancement layers will also be delivered 
without packet loss or delay but when congestion occurs we demote the enhancement layers to best 
effort traffic and prioritize the base layer. Research discussed in [4] also attempts to route 
the enhancement layers as a lossy-QoS flow. Our research goes further and also enables fast failure 
recovery when a link fails.\\

\section{Architecture}
Each OpenFlow network contains at least one controller which is responsible for managing all the switches. 
A default OpenFlow switch forwards packets 
to the controller if the packet does not match any forwarding entry. 
We route SVC encoded video streams as a QoS flow, where we consider the base layer as a 
high priority QoS flow and the enhancement layer as either a lower QoS flow or as best effort traffic. 
These principals are shown in Figure \ref{fig_sim}. A filter on the controller separates QoS flows from 
other traffic, after which the controller can choose a path for the different layers. 
When the controller calculates a path for a layer, the new forwarding rules are sent to the switches.\\

\begin{figure*}[htb]
\centering
\includegraphics[width=\linewidth]{fig/SVC_OpenFlow_flow}
\caption{Simulation Results}
\label{fig_sim}
\end{figure*}

We defined a cost function which can be used to calculate the cost of a path.
The cost function is based on the following observations. (Insert ref here)
\begin{itemize}
	\item When high priority QoS flows are forwarded on the same link as best effort traffic, there will be packet loss in the best effort flow to prioritize the QoS flow.
	\item Metrics for QoS flows must be different than those used for best effort traffic. The shortest path and QoS path must be different to obtain good results. 
	If both follow the same path, the best effort traffic will experience more packet loss.
	\item Calculating a path for a new QoS flow does not change the best effort routing tables. The packet loss of best effort traffic caused by QoS flows must be considered during path selection.
	\item The length of alternative paths for QoS flows should still be kept as short as possible to optimize transfer time.
	\item Best effort traffic can be used as a reference to retrieve metrics from a switch or link, such as packet loss or to estimate the amount of traffic on that route.
\end{itemize}

These observations show that packet loss and path length are part of our cost function: 
the cost function must find a balance between maintaining low packet loss of the best effort 
traffic and the delay of QoS flows, caused by using longer paths.\\

Furthermore, the controller 
must be able to obtain these metrics at any time from any switch in the network before calculating 
new flow tables. We implemented a monitor module on the controller to gather all network 
statistics. Besides polling for statistics, a switch must also forward events to the controller when 
they occur, such as an link going down or a flow experiencing packet loss. These events provide 
additional monitoring data for the routing algorithms on the controller.\\

\section{QoS Routing}
Must ensure zero packet loss base layer, optimize other traffic ... cost function?
Go from theoretical cost function to exact function in code?
--> Yes

\section{Evaluation}
Test scenario (client/server - filter based on port) + metrics for result + present results\\
An evaluation metric has to represent a value which is relevant to what we want to improve, which is why a measurement must:
\begin{itemize}
	\item represent the quality of the received video
	\item represent the performance of the network
	\item represent the impact on best effort traffic
\end{itemize}
To measure quality of the video we have chosen for the PSNR value of the received video compared to the raw footage. The higher the PSNR value, the better the frames were preserved during transfer. To measure the performance of the network we used the bitrate of all traffic. We also show the packet loss of best effort traffic caused by prioritizing the QoS flow...
